---
title: "Symbiodiniaceae-bacteria interactions and functions supporting the formation of calcifying biofilms"
author: "MNitschke et al"
date: "100819"
output:
  pdf_document: default
  html_document: default
  keep_md : TRUE
---

R scripts for reproducing the analysis of the microbiomes of Symbiodinium cultures. These scripts follow on from the Qiime + Vsearch pipeline used to generate OTUs and assign taxonomy against the Silva 132 database.The primary input is an OTU table in biom format that already contains the sample and observation metadata ("otu_table132.taxfinalmeta.biom").

# Libraries

```{r}
library(phyloseq)
library(phyloseq.extended)
library(DESeq2)
library(vegan)
library(ggplot2)
library(dplyr)
library(stringr)
library(scales)
library(data.table)
library(car)
library(FSA)
library(genefilter)
```

Note: loading in phyloseq.extended from https://github.com/mahendra-mariadassou/phyloseq-extended to produce rarefaction curves.

# Import data 

Grab the .biom file which is the final step of the Qiime1 + VSearch pipeline.

```{r}
biom <- import_biom("otu_table132.taxfinalmeta.biom")

# Add ITS2 type to sample metadata
ITS2 <- c("A2","A2","B1","B1","A2","A2","A2","A2","A2","A2","A2","A2","A2","B1","A1","A1","A1","A1","A1","A1","A1","A1","A1","A4","B1","A2","B1","A12","A2","A2","A2","A2","A2","A2","A2","A2","A2","B1","B1","B1","B1","B1")

# Clade = genus (A, Symbiodinium; B, Breviolum)
clade <- c("A","A","B","B","A","A","A","A","A","A","A","A","A","B","A","A","A","A","A","A","A","A","A","A","B","A","B","A","A","A","A","A","A","A","A","A","A","B","B","B","B","B")

sample_data(biom)$ITS2 <- ITS2
sample_data(biom)$clade <- clade
```

# Supplementary Figure S1: rarefaction curves

```{r}
p <- phyloseq.extended:::ggrare(biom, step = 100, color = "state", label = "Sample", se = FALSE) + 
  theme(aspect.ratio = 9 / 16)
```



# SP vs NP

Subset the dataset by removing the temporal pre-calcifying and post-calcifying data. 

```{r}
SPNP <- subset_samples(biom, time=="n")
```

## Alpha diversity
### Supplementary Table S2

Multiple rarify (100 replications) to minimum read depth across samples to compute alpha diversity statistics (richness, diversity)

```{r}
min_lib <- min(sample_sums(SPNP))
nsamp = nsamples(SPNP)
trials = 100
richness <- matrix(nrow = nsamp, ncol = trials)
row.names(richness) <- sample_names(SPNP)
diversity <- matrix(nrow = nsamp, ncol = trials)
row.names(diversity) <- sample_names(SPNP)
set.seed(3) # Set a seed so results are reproducible

for (i in 1:100) {
   
   r <- rarefy_even_depth(SPNP, sample.size = min_lib, verbose = FALSE, replace = TRUE)
   
   rich <- as.numeric(as.matrix(estimate_richness(r, measures = "Observed")))
   richness[ ,i] <- rich
   
   diver <- as.numeric(as.matrix(estimate_richness(r, measures = "InvSimpson")))
   diversity[ ,i] <- diver
}

SampleID <- row.names(richness)
Sampledata <- sample_data(SPNP)$state
mean <- apply(richness, 1, mean)
sd <- apply(richness, 1, sd)
measure <- rep("Richness", nsamp)
rich_stats <- data.frame(SampleID, mean, sd, Sampledata, measure)
rich_stats

SampleID2 <- row.names(diversity)
Sampledata2 <- sample_data(SPNP)$state
mean2 <- apply(diversity, 1, mean)
sd2 <- apply(diversity, 1, sd)
measure2 <- rep("diversity", nsamp)
diversity_stats <- data.frame(SampleID2, mean2, sd2, Sampledata2, measure2)
diversity_stats

write.table(rich_stats, "rich.tsv", sep="\t", row.names = TRUE, col.names = TRUE, quote = FALSE)
write.table(diversity_stats, "diverse.tsv", sep="\t", row.names = TRUE, col.names = TRUE, quote = FALSE)
```

## Alpha diversity statistics

Order the table and conduct t test on calc vs non calc

```{r}
kruskal.test(mean ~ Sampledata, data = rich_stats)
kruskal.test(mean2 ~ Sampledata2, data = diversity_stats)

t.test(mean ~ Sampledata, data = rich_stats)
t.test(mean2 ~ Sampledata2, data = diversity_stats)
```

## Figure 1a: Unconstrained ordination

SP vs NP ordination using a Bray-Curtis matrix and PCoA ordination.

```{r}
calctrans <- transform_sample_counts(SPNP, function(x) x/sum(x))
brayordtrans <- ordinate(calctrans, "PCoA", "bray")
plot_ordination(calctrans, brayordtrans, color="state", shape="clade")  + 
  geom_text(aes(label = sample_names(calctrans)), vjust = 1.5) + 
  geom_point(size = 4)
```

## Figure 1a part 2

Grab centroids and vectors from samples to centroids to overlay on part 1

```{r}
bray <- phyloseq::distance(calctrans, method = "bray")

sampledf <- data.frame(sample_data(calctrans))

betastate <- betadisper(bray, sampledf$state)

plot(betastate)
```

## PERMANOVA and betadisper

Run script chunk above first and then compute statistics and homogeneity of dispersions test

## Supplementary Table S1
```{r}
adonis2(bray ~ state + clade, by = "margin", data = sampledf)

anova(betastate)
```

## Heatmap: Supplementary Figure S3

Create order so sister strains are adjacent, and also so that SP and NP strains are grouped.

```{r}
order <- as.character(c("107","108","64","RT64","13","RT13","1852A","1852B","1852C","3622A","3622B","3622C","89","MF2A","MF2B","MF2C","RT146","146","351","RT351","379","Aip24B","CX","MacPC"))
```

Summarize by genus-level taxonomy assignments, transform to relative proportions, and plot heatmap using a bray-curtis matrix and the PCoA method.

```{r}
gen_glomcalc <- tax_glom(SPNP, taxrank="Rank6")
gen_glomcalc = transform_sample_counts(gen_glomcalc, function(x) x/sum(x))

TOP=names(sort(taxa_sums(gen_glomcalc),TRUE)[1:25])
TOP=prune_taxa(TOP,gen_glomcalc)

plot_heatmap(TOP, "PCoA", "bray", sample.order = order, low="#000033", high="#FF3300", taxa.label = "Rank6", trans=identity_trans())

rare=names(sort(taxa_sums(gen_glomcalc),TRUE)[20:66])
rare=prune_taxa(rare,gen_glomcalc)

plot_heatmap(rare, "PCoA", "bray", sample.order = order, low="#000033", high="#338DF4", taxa.label = "Rank6", trans = identity_trans())
```

## Differential abundance

### Supplementary table S3

Use DESeq2 to test for differential abundance of phylum-level assignments between SP and NP strains.

```{r}
phy_glom <- tax_glom(SPNP, taxrank = "Rank2")
SPvNP.phy <- phyloseq_to_deseq2(phy_glom, ~ state)
SPvNP.phy = DESeq(SPvNP.phy, test="Wald", fitType="parametric")
res = results(SPvNP.phy, cooksCutoff = FALSE, contrast = c("state","SP","NP"))
alpha = 0.05
sigtabp = res[which(res$padj < alpha), ]
sigtabp = cbind(as(sigtabp, "data.frame"), as(tax_table(phy_glom)[rownames(sigtabp), ], "matrix"))
View(sigtabp)
```

Use DESeq2 to test for differential abundance of family-level assignments between SP and NP strains.

```{r}
fam_glom <- tax_glom(SPNP, taxrank="Rank5")
SPvNP.gen <- phyloseq_to_deseq2(fam_glom, ~ state)
SPvNP.gen = DESeq(SPvNP.gen, test="Wald", fitType="parametric")
res = results(SPvNP.gen, cooksCutoff = FALSE)
alpha = 0.05
sigtabf = res[which(res$padj < alpha), ]
sigtabf = cbind(as(sigtabf, "data.frame"), as(tax_table(fam_glom)[rownames(sigtabf), ], "matrix"))
View(sigtabf)
```

Use DESeq2 to test for differential abundance of genus-level assignments between SP and NP strains.

```{r}
gen_glom <- tax_glom(SPNP, taxrank="Rank6")
SPvNP.gen <- phyloseq_to_deseq2(gen_glom, ~ state)
SPvNP.gen = DESeq(SPvNP.gen, test="Wald", fitType="parametric")
res = results(SPvNP.gen, cooksCutoff = FALSE)
alpha = 0.05
sigtabg = res[which(res$padj < alpha), ]
sigtabg = cbind(as(sigtabg, "data.frame"), as(tax_table(gen_glom)[rownames(sigtabg), ], "matrix"))
View(sigtabg)
```

### Figure 1b Differential Families

```{r}
theme_set(theme_bw())
scale_fill_discrete <- function(palname = "Set1", ...) {
    scale_fill_brewer(palette = palname, ...)
}

x = tapply(sigtabf$log2FoldChange, sigtabf$Rank5, function(x) max(x))
x = sort(x, TRUE)

error <- sigtabf$lfcSE

sigtabf$Rank5 = factor(as.character(sigtabf$Rank5), levels=names(x))

p <- ggplot(sigtabf, aes(x = log2FoldChange, y = Rank5, color = Rank5)) + 
                    geom_errorbarh(aes(xmin = log2FoldChange + error, xmax = log2FoldChange - error), height = 0.1, colour = "black") + 
                    geom_point(size=5) + 
                    theme(axis.text.x = element_text(angle = 90, hjust = 0, vjust=0.5)) +
                    theme(aspect.ratio = 16 / 9)
```

## Supplementary Figure S2: Phylum level relative proportions

```{r}
phy_glom <- tax_glom(SPNP, taxrank="Rank2")
phy_glom = transform_sample_counts(phy_glom, function(x) x/sum(x))

plot_bar(phy_glom, fill = "Rank2") + 
  facet_grid(~state, scales = "free", space = "free_x")
```

## Supplementary Table S4: Core microbiome

Compute core to compare to Lawson et al 2017. Filter dataset to remove an OTU if it is not present in all samples.

```{r}
n <- length(sample_names(biom))
flist <- filterfun(kOverA(n, 0))
a <- filter_taxa(biom, flist)
core = prune_taxa(a, biom)
coreglom <- tax_glom(core, taxrank = "Rank6")
```

```{r}
coretab <- as.data.frame(tax_table(coreglom))
coretab$Rank7 <- NULL
write.table(coretab,"core.tsv", sep="\t", row.names = TRUE, col.names = TRUE, quote = FALSE)
```

# SP cultures through time

Subset to only the data that has pre, active, and post calcifying phases (185, Mf, 362). 

Note: In the sample metadata culture "a" = strain 185. "b" = strain 362. "c" = strain Mfav.

```{r}
subtemporal <- subset_samples(biom, series=="y")
```

## Alpha diversity

Multiple rarify (100 replications) to minimum read depth across samples to compute alpha diversity statistics (richness, diversity)

```{r}
min_lib.temp <- min(sample_sums(subtemporal))
nsamp.temp = nsamples(subtemporal)
trials = 100
richness.temp <- matrix(nrow = nsamp.temp, ncol = trials)
row.names(richness.temp) <- sample_names(subtemporal)
diversity.temp <- matrix(nrow = nsamp.temp, ncol = trials)
row.names(diversity.temp) <- sample_names(subtemporal)
set.seed(3)

for (i in 1:100) {
   
   r.temp <- rarefy_even_depth(subtemporal, sample.size = min_lib.temp, verbose = FALSE, replace = TRUE)
   
   rich.temp <- as.numeric(as.matrix(estimate_richness(r.temp, measures = "Observed")))
   richness.temp[ ,i] <- rich.temp
   
   diver.temp <- as.numeric(as.matrix(estimate_richness(r.temp, measures = "InvSimpson")))
   diversity.temp[ ,i] <- diver.temp
}

SampleID.temp <- row.names(richness.temp)
Sampledata.temp <- sample_data(subtemporal)$phase
Samplestrain <- sample_data(subtemporal)$culture
mean.temp <- apply(richness.temp, 1, mean)
sd.temp <- apply(richness.temp, 1, sd)
measure1.temp <- rep("Richness", nsamp.temp)
rich_stats.temp <- data.frame(SampleID.temp, mean.temp, sd.temp, Sampledata.temp, Samplestrain, measure1.temp)

rich_stats.temp

mean2.temp <- apply(diversity.temp, 1, mean)
sd2.temp <- apply(diversity.temp, 1, sd)
measure2.temp <- rep("diversity", nsamp.temp)
diversity_stats.temp <- data.frame(SampleID.temp, mean2.temp, sd2.temp, Sampledata.temp, Samplestrain, measure2.temp)

diversity_stats.temp

write.table(rich_stats.temp, "rich_temporal.tsv", sep="\t", row.names = TRUE, col.names = TRUE, quote = FALSE)
write.table(diversity_stats.temp, "diverese_temporal.tsv", sep="\t", row.names = TRUE, col.names = TRUE, quote = FALSE)
```

### Alpha statistics

Run above chunk of code, collect the alpha diversity metrics for each strain, and perform non-parametric multiple comparisons to look for differences. Note: a = strain 185, b = strain 362, c = strain M.fav

```{r}
ar <- rich_stats.temp[rich_stats.temp$Samplestrain == 'a',]
br <- rich_stats.temp[rich_stats.temp$Samplestrain == 'b',]
cr <- rich_stats.temp[rich_stats.temp$Samplestrain == 'c',]

dunnTest(mean.temp ~ Sampledata.temp, data = ar)
dunnTest(mean.temp ~ Sampledata.temp, data = br)
dunnTest(mean.temp ~ Sampledata.temp, data = cr)

ae <- diversity_stats.temp[diversity_stats.temp$Samplestrain == 'a',]
be <- diversity_stats.temp[diversity_stats.temp$Samplestrain == 'b',]
ce <- diversity_stats.temp[diversity_stats.temp$Samplestrain == 'c',]

dunnTest(mean2.temp ~ Sampledata.temp, data = ae)
dunnTest(mean2.temp ~ Sampledata.temp, data = be)
dunnTest(mean2.temp ~ Sampledata.temp, data = ce)
```

### Supplementary Figure S5: Alpha diversity bar plots 

Plot differences in alpha diversity across all phases

```{r}
positions <- c("pre", "active", "post")

Richplot <- aggregate(rich_stats.temp[, 2:3], list(rich_stats.temp$Sampledata.temp, rich_stats.temp$Samplestrain), mean)

rp <- ggplot(data=Richplot, aes(x=Group.1, y=mean.temp)) + 
  geom_bar(stat = "identity") + facet_grid(~Group.2) + 
  geom_errorbar(aes(ymin=mean.temp-sd.temp/sqrt(3), ymax=mean.temp+sd.temp/sqrt(3)), width=0.2) +
  scale_x_discrete(limits = positions) +
  theme(aspect.ratio = 1)
  
rp

diverplot <- aggregate(diversity_stats.temp[, 2:3], list(diversity_stats.temp$Sampledata.temp, diversity_stats.temp$Samplestrain), mean)

dp <- ggplot(data=diverplot, aes(x=Group.1, y=mean2.temp)) + 
  geom_bar(stat = "identity") + facet_grid(~Group.2) + 
  geom_errorbar(aes(ymin=mean2.temp-sd2.temp/sqrt(3), ymax=mean2.temp+sd2.temp/sqrt(3)), width=0.2) +
  scale_x_discrete(limits = positions) +
  theme(aspect.ratio = 1)

dp
```

## Figure 3a: Unconstrained ordination

Transform to relative proportions as above. Pre vs active vs post calcification ordinations using a Bray-Curtis matrix and PCoA ordination.

```{r}
subtemptrans = transform_sample_counts(subtemporal, function(x) x/sum(x))

brayordtranstemp <- ordinate(subtemptrans, "PCoA", "bray")

plot_ordination(subtemptrans, brayordtranstemp, color="phase", shape="phase")  + 
  geom_text(aes(label = sample_names(subtemptrans)), vjust = 1.5) + 
  geom_point(size = 4)
```

## Figure 3a part 2 

Grab centroids and vectors from samples to centroids to overlay on part 1

```{r}
subtemptransbray <- phyloseq::distance(subtemptrans, method = "bray")

df <- data.frame(sample_data(subtemptrans))

beta2 <- betadisper(subtemptransbray, df$phase)

plot(beta2)
```

## PERMANOVA and betadisper

### Supplementary table S7

```{r echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
adonis2(formula = subtemptransbray ~ phase, data = df)
# First model (phase only) does not detect differences. Constrain permutations to be nested within each culture to detect differences in phase specific to each culture.

adonis2(formula = subtemptransbray ~ culture/phase, data = df, strata = df$culture)

permutest(beta2)
```

## Differential abundance

Use DESeq2 to look for common genera that change through time across all three SP strains sampled in series. Note: Positive Log2 foldchange values represent higher abundance in pre vs active (first test), and then higher abundance in active vs post (second test). Negative values represent the opposite.

```{r}
subtemporal <- subset_samples(biom, series=="y")

phasegen <- tax_glom(subtemporal, taxrank="Rank6")

# Start by treating phases as replicates and ignore cultures
phasegend2 <- phyloseq_to_deseq2(phasegen, ~ phase)

phasegend2 = DESeq(phasegend2, test="Wald", fitType="parametric")

res.p.a = results(phasegend2, cooksCutoff = FALSE, contrast = c("phase", "pre", "active"))
res.a.po = results(phasegend2, cooksCutoff = FALSE, contrast = c("phase", "active", "post"))
res.p.po = results(phasegend2, cooksCutoff = FALSE, contrast = c("phase", "pre", "post"))

alpha = 0.05

sigtab.p.a = res.p.a[which(res.p.a$padj < alpha), ]
sigtab.p.a = cbind(as(sigtab.p.a, "data.frame"), as(tax_table(phasegen)[rownames(sigtab.p.a), ], "matrix"))

View(sigtab.p.a)

sigtab.a.po = res.a.po[which(res.a.po$padj < alpha), ]
sigtab.a.po = cbind(as(sigtab.a.po, "data.frame"), as(tax_table(phasegen)[rownames(sigtab.a.po), ], "matrix"))

View(sigtab.a.po)

sigtab.p.po = res.p.po[which(res.p.po$padj < alpha), ]
sigtab.p.po = cbind(as(sigtab.p.po, "data.frame"), as(tax_table(phasegen)[rownames(sigtab.p.po), ], "matrix"))

View(sigtab.p.po)

# Next explore across cultures

phasegend2 <- phyloseq_to_deseq2(phasegen, ~ culture * phase) 

phasegend2 = DESeq(phasegend2, test="Wald", fitType="parametric")

res.p.a = results(phasegend2, cooksCutoff = FALSE, contrast = c("phase", "pre", "active"))
res.a.po = results(phasegend2, cooksCutoff = FALSE, contrast = c("phase", "active", "post"))
res.p.po = results(phasegend2, cooksCutoff = FALSE, contrast = c("phase", "pre", "post"))

alpha = 0.05

sigtab.p.a = res.p.a[which(res.p.a$padj < alpha), ]
sigtab.p.a = cbind(as(sigtab.p.a, "data.frame"), as(tax_table(phasegen)[rownames(sigtab.p.a), ], "matrix"))

View(sigtab.p.a)

sigtab.a.po = res.a.po[which(res.a.po$padj < alpha), ]
sigtab.a.po = cbind(as(sigtab.a.po, "data.frame"), as(tax_table(phasegen)[rownames(sigtab.a.po), ], "matrix"))

View(sigtab.a.po)

sigtab.p.po = res.p.po[which(res.p.po$padj < alpha), ]
sigtab.p.po = cbind(as(sigtab.p.po, "data.frame"), as(tax_table(phasegen)[rownames(sigtab.p.po), ], "matrix"))

View(sigtab.p.po)
```

### Figure 3b

Plotting differentially abundant across phases ignoring culture as factor in design

```{r}
positions <- c("pre", "active", "post")

tr <- transform_sample_counts(subtemporal, function(x) x/sum(x)*100)

df <- psmelt(tr)

df$Rank7 <- NULL

avgs <- df %>% select(OTU, Sample, culture, Abundance, Rank6, phase)

d = avgs %>%
    group_by(Sample,culture,Rank6,phase) %>%
    filter(Abundance > 0) %>%
    summarize(rank6.abund = sum(Abundance))

d2 = d %>%
    group_by(Rank6,phase) %>%
    #filter(Abundance > 0) %>%
    summarize(abund = mean(rank6.abund), se = sd(rank6.abund)/sqrt(3))

# Use str_detect to select only the differentially abundant genera across all cultures
d2$Rank6 <- as.factor(d2$Rank6)
d2$Rank6 <- factor(d2$Rank6, levels = c("Marinobacter","SM1A02","Nitratireductor","BD1-7 clade")) 
d2$phase <- as.factor(d2$phase)

d2 <- d2 %>% filter(str_detect(Rank6, "Marinobacter|SM1A02|Nitratireductor|BD1-7 clade"))

dp <- ggplot(data=d2, aes(x=phase, y=abund)) + 
  geom_bar(stat = "identity") + facet_grid(Rank6~., scales = "free_y") + 
  geom_errorbar(aes(ymin=abund-se, ymax=abund+se), width=0.1) + 
  scale_x_discrete(limits = positions) + 
  scale_y_continuous() + 
  theme(aspect.ratio = 1)

dp
```

### Supplementary Figure S6:

Plotting differentially abundant across phases including culture as factor in design.

```{r}
positions <- c("pre", "active", "post")

tr <- transform_sample_counts(subtemporal, function(x) x/sum(x)*100)

df <- psmelt(tr)

df$Rank7 <- NULL

avgs <- df %>% select(OTU, Sample, culture, Abundance, Rank6, phase)

d = avgs %>%
    group_by(Sample,culture,Rank6,phase) %>%
    filter(Abundance > 0) %>%
    summarize(rank6.abund = sum(Abundance))

d2 = d %>%
    group_by(culture,Rank6,phase) %>%
    summarize(abund = mean(rank6.abund), se = sd(rank6.abund)/sqrt(3))

# Use str_detect to select differentially abundant genera detected when including culture as a parameter in the deseq2 model
d2 <- d2 %>% filter(str_detect(Rank6,"SM1A02|OM60|Owenweeksia|Ekhidna|Balneola|Mf105b01|Marinobacter|Oceanicaulis|Labrenzia|Pseudohongiella"))

d2$culture <- as.factor(d2$culture)
d2$Rank6 <- as.factor(d2$Rank6)
d2$Rank6 <- factor(d2$Rank6, levels = c("Labrenzia","Marinobacter","SM1A02","Balneola","OM60(NOR5) clade","Mf105b01","Oceanicaulis","Owenweeksia","Pseudohongiella","Ekhidna"))
d2$phase <- as.factor(d2$phase)

dp <- ggplot(data=d2, aes(x=phase, y=abund)) + 
  geom_bar(stat = "identity", aes(fill = phase)) + 
  facet_grid(Rank6 ~ culture, scales = "free_y") + 
  geom_errorbar(aes(ymin=abund-se, ymax=abund+se), width=0.1) + 
  scale_x_discrete(limits = positions) + scale_y_continuous() + 
  theme(aspect.ratio = 1)

dp
```
