---
title: "Xeq and Chl: Symbiodinium-bacteria interactions support the formation of calcifying biofilms""
author: "MNitschke"
date: "100819"
output: html_document
---

# Libraries

```{r}
library(readr)
library(tidyr)
library(stringr)
library(ggplot2)
library(scales)
library(dplyr)
library(broom)
library(FSA)
```

# Import acidic polysacharride and Chl data

```{r}
XC <- read.table(file = 'XeqChl.txt', sep = '\t', header = TRUE)
```

# Figure 5 part 2

Plot

```{r}
XCsis <- filter(XC, sister != "n")
XCsis <- filter(XCsis, sister != "sp64")

p <- ggplot(data = XCsis, aes(x = time, group = identity)) +
      geom_line(aes(y = chl)) +
      geom_line(aes(y = xeq/200)) +
      geom_point(aes(y = chl, colour = "chl", shape = identity), size = 3.5) +
      geom_point(aes(y = xeq/200, colour = "xeq", shape = identity), size = 3.5) +
      geom_errorbar(mapping = aes(x = time, y = chl, ymin = chl - chlse, ymax = chl + chlse), width = 0.2) +
      geom_errorbar(mapping = aes(x = time, y = xeq/200, ymin = (xeq/200) - (xeqse/200), ymax = (xeq/200) + (xeqse/200)), width = 0.2) +
      scale_y_continuous(sec.axis = sec_axis(~.*200, name = "Xeq")) +
      scale_color_manual(values = c(chl = "#00F484", xeq = "#35A2FF")) +
      facet_wrap(~sister, nrow = 3, ncol = 9) +
      theme(aspect.ratio = 1, panel.background = element_rect(fill = "white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank())
      
p
```

# Supplementary figure S7

Chl a OD (absorbance in vivo 675 nm) and acidic polysacharrides (Xeq ug) through time for SP and NP cultures.

```{r}
XC$order <- factor(XC$culture, levels = unique(XC$culture))

p <- ggplot(data = XC, aes(x = time)) +
      geom_line(aes(y = chl)) +
      geom_line(aes(y = xeq/200)) +
      geom_point(aes(y = chl, colour = "chl"),size = 2) +
      geom_point(aes(y = xeq/200, colour = "xeq"),size = 2) +
      geom_errorbar(mapping = aes(x = time, y = chl, ymin = chl - chlse, ymax = chl + chlse), width = 0.1) +
      geom_errorbar(mapping = aes(x = time, y = xeq/200, ymin = (xeq/200) - (xeqse/200), ymax = (xeq/200) + (xeqse/200)), width = 0.1) +
      scale_y_continuous(sec.axis = sec_axis(~.*200, name = "Xeq")) +
      scale_color_manual(values = c(chl = "#00F484", xeq = "#35A2FF")) +
      facet_wrap(~order, nrow = 5, ncol = 6) +
      theme(aspect.ratio = 1, panel.background = element_rect(fill = "white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank())
      
p
```

# Linear model Acidic polysacharrides (ug Xeq) vs Chl (OD 675 nm)

## Suplementary tables S11 and S12

```{r}
fitlm = XC %>% group_by(culture) %>% do(model = lm(xeq ~ chl, data = .))
fitlm %>% tidy(model)
fitlm %>% glance(model)
fitlm %>% augment(model)
```

# Supplementary figure S9

Chl a OD (absorbance in vivo 675 nm) vs acidic polysacharrides (Xeq ug) through time for SP and NP cultures.

```{r}
theme_set(theme_bw())
p <- ggplot(data=XC, aes(x = chl, y=xeq, sample = culture)) +
      geom_smooth(method = "lm", se=TRUE, color="red", formula = y ~ x) +
      geom_path() +
      geom_errorbar(mapping = aes(x = chl, y = xeq, ymin = xeq - xeqse, ymax = xeq + xeqse),width = 0.01) +
      geom_errorbarh(mapping = aes(x = chl, y = xeq, xmin = chl - chlse, xmax = chl + chlse, height = 2)) +
      geom_point(aes(x = chl, y = xeq), size = 2, shape = 21, fill = "white") +
      facet_wrap(~order, nrow = 5, ncol = 6) +
      theme(aspect.ratio = 1)
      
p
```

# Import bacterial biofilm and acidic polysacharride data

```{r}
XB <- read.table(file = 'xeqbiofilm.txt', sep = '\t', header = TRUE)
```

# Xeq vs biofilm

## Figure 6

```{r}
sel <- XB %>% select(genus, species, strain, xeq, biofilm)

XBmse = sel %>%
    group_by(genus,species,strain) %>%
    summarize(xeq.m = mean(xeq), xeq.se = sd(xeq)/sqrt(6), biofilm.m = mean(biofilm), biofilm.se = sd(biofilm)/sqrt(6))

p <- ggplot(data=XBmse, aes(x=biofilm.m, y=xeq.m)) +
      geom_errorbar(mapping = aes(x = biofilm.m, y = xeq.m, ymin = xeq.m-xeq.se, ymax = xeq.m + xeq.se),width = 0.01) +
      geom_errorbarh(mapping = aes(x = biofilm.m, y = xeq.m, xmin=biofilm.m-biofilm.se, xmax=biofilm.m+biofilm.se),height=0.8) +
      geom_point(aes(colour = species), size = 6) +
      geom_text(aes(label = strain), vjust = 0, nudge_y = -5) +
      theme(aspect.ratio = 1, panel.background = element_rect(fill = "white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank())

p
```

# Supplementary methods figure SM1: calibration curve

Standard curve of Xanthan Gum (?g) vs optical density (absorbance 787nm) of acidified (80% H2SO4) alcian blue following Claquin et al (2008) Aquatic Microbial Ecology.

```{r}
fit <- lm(a787 ~ Xug, data = XC)

summary(fit)

p <- ggplot(data=XC, aes(x=Xug, y=a787)) +
      geom_point(size = 2) +
      geom_smooth(method = "lm", se=TRUE, color="red", formula = y ~ x) +
      theme(aspect.ratio = 1, panel.background = element_rect(fill = "white"), panel.grid.major = element_blank(), panel.grid.minor = element_blank())

p
```

